### 2025-08-26 sections/5-discussion.tex
tightening discussion

We have presented CAKES, a set of three algorithms for fast $k$-NN search that are generic over a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (see definition in Section~\ref{sec:methods}).
Even under cosine distance, which is not a metric, CAKES's algorithms exhibit nearly perfect recall.
CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
CAKES uses an unbalanced clustering algorithm to build a binary tree that better captures the manifold structure of the data and provides for faster search performance.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.

@@

As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on real-world datasets with low LFD, and scale linearly on synthetic datasets with high LFD.
CAKES's performance on Sift against that on the Random dataset illustrates this phenomenon well.
As seen in Figures~\ref{fig:results:random-lfd}~and~\ref{fig:results:sift-lfd}, the LFD of the Random dataset is much higher than that of Sift, even though both datasets have the same cardinality and dimensionality.
This is unsurprising, given that the Random dataset is uniformly distributed, while Sift is not.
On the Random dataset, CAKES's speed decreases linearly as the multiplier increases, while with Sift, Depth-First Sieve and Breadth-First Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit near-constant throughput with Sift, throughput decreases much more slowly than it does with the Random dataset.
Since these two datasets have the same cardinality and embedding dimension, the aforementioned discrepancies highlight how manifold structure affects algorithm performance.

We emphasize that although CAKES's algorithms are slower on the Random dataset, their recalls remain perfect.
Figure~\ref{fig:results:scaling-plots} also shows that HNSW and ANNOY have near-constant throughput as the multiplier increases, but their recall continues to decrease as the multiplier increases.
This, coupled with the increasing time and space cost of building the indices for HNSW and ANNOY, makes these algorithms unsuitable for real-world applications in which the datasets are growing exponentially.
CAKES's tree is orders of magnitude faster and less memory-intensive to build, and CAKES's algorithms, while slower than HNSW and ANNOY, still show near-constant scaling as well as perfect recall.

@@

While the fastest CAKES algorithm can vary by dataset and cardinality, we note that on all three ANN-benchmark datasets (Fashion-MNIST, Glove-25, and Sift), Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as the cardinality increases.
This observation warrants further investigation, but it is especially promising given that both algorithms consistently outperform linear search for high cardinalities.
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities supports our use of an auto-tuning function (as discussed in Section~\ref{sec:methods:auto-tuning}) to select the best algorithm for a given dataset.
Future work is needed to better characterize those datasets for which each algorithm performs best, but our results suggest that a more sophisticated auto-tuning function could be developed to select the fastest algorithm based on the properties of the dataset.

We also stress that CAKES's algorithms are designed to work well for \textit{big} data, and our results support this claim;
we observe that while linear search can outperform CAKES on datasets with small cardinality, CAKES always overtakes linear search at a large enough cardinality for each of the ANN-benchmark datasets we tested.
Moreover, we observe that CAKES's algorithms outperform linear search at a lower cardinality when the dataset has a lower LFD.
For example, Sift has much higher LFD than Fashion-Mnist and Glove-25, and we observe that CAKES overtakes linear search at a cardinality of $10^7$ for Sift, as opposed to about $10^5$ and $10^6$ for Fashion-Mnist and Glove-25, respectively.
For the Random dataset, which has the highest LFD, CAKES's algorithms \textit{never} outperform linear search.
These observations support our claim that CAKES performs well on datasets that arise from constrained generating phenomena, even as the cardinalities of these datasets grow exponentially.

@@

We note that despite the genericity of CAKES, effectively applying it to real-world datasets may require domain-specific knowledge.
As an example, consider the domain of protein sequence search; protein sequences are of highly variable length (30--1000 amino acids are typical, with some outliers) and typically comprise a 20-letter alphabet.
There have been numerous protein sequence search algorithms~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and the one thing they have in common besides the application domain is that their development required deep domain-specific knowledge.
While simple edit distance may be a poor choice for clustering when sequence lengths exhibit such variance, other choices of distance metrics and preprocessing could be used by CAKES, such as representing each sequence as a set of $k$-mers (short substrings of length $k$) and choosing Jaccard distance on these sets as the metric; this approach was chosen by~\cite{kim2021entrance}.
If the distance function uses a global sequence alignment algorithm, such as Needleman-Wunsch~\cite{needleman1970general}, the dataset may also be preprocessed into a number of mutually exclusive subsets based on the lengths of the sequences, and a separate CAKES tree may be built for each subset.
At search time, the query sequence would be assigned to the appropriate tree based on its length, and the search would be performed on that tree, and on a small number of trees on adjacent lengths' bins.

@@

The questions raised by this study suggest several additional avenues for future work.
A comparison across more datasets is in order, as is further analysis with other distance functions that existing methods, such as FAISS, HNSW and ANNOY, do not support.
These include Wasserstein distance~\cite{vallender1974calculation} for probability distributions (particularly for high dimensional distributions) and Tanimoto distance~\cite{bajusz2015tanimoto} for comparing molecular structures by their maximal common subgraphs.
Incorporating new distance functions in CAKES requires only that a Rust implementation of the distance function be provided.

Additionally, we intend to improve the data augmentation process described in Section~\ref{sec:methods:synthetic-data} to better preserve the topological structure of the underlying dataset.
In particular, we expect that favoring augmentation along the first few principal components of the local manifold may allow for more realistic data augmentation.

% We already did compression so it feels weird to have this in there:

% We also plan to investigate hierarchical data compression by representing differences at each level of the binary tree, particularly for string or genomic data, where all differences are discrete.
% Such an encoding would allow us to store the data in a compressed format, and to perform search on the compressed data without decompressing it in its entirety.
% This would allow us to perform fast search on datasets that are otherwise too large to fit in memory.
% An exploration of compression and compressed search is another avenue for future work.

We also would like to explore the use of CAKES in a streaming environment.
This would require the ability to perform ``online'' updates to the tree as points are added to or deleted from the dataset.
Such online updates would take advantage of the fast search algorithms provided by CAKES.  % and would allow us to have our cake and eat it, too.
CAKES can also be used to extend anomaly detection in CHAODA~\cite{ishaq2021clustered}.
We could add to CHAODA's ensemble of graph-based anomaly detection methods by using the distribution of distances among the $k$ nearest neighbors of cluster centers.

@@

CLAM and CAKES are implemented in the Rust programming language, and the source code is available under an MIT license at https://github.com/URI-ABD/clam.

>>>>

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions. When the distance is a metric (Sec.~\ref{sec:methods}), the algorithms are exact; under cosine distance (non-metric) they achieve nearly perfect recall. CAKES is most effective when data satisfy the manifold hypothesis—lying on a low-dimensional manifold within a high-dimensional space. It builds an unbalanced-clustering binary tree that captures manifold structure and accelerates search. Conversely, performance degrades on randomly distributed data, where such structure is absent.

@@

As expected, Fig.~\ref{fig:results:scaling-plots} shows CAKES scaling sublinearly with cardinality on real-world datasets with low LFD and linearly on synthetic datasets with high LFD; Sift versus Random illustrates this contrast.

On Random, CAKES throughput decreases linearly with the multiplier, whereas on Sift, Depth-First Sieve and Breadth-First Sieve are nearly constant. Repeated $\rho$-NN is not near-constant on Sift but decays much more slowly than on Random. Because Sift and Random share cardinality and embedding dimension, this isolates the effect of manifold structure on performance.

Although CAKES is slower on Random, its recall remains perfect. HNSW and ANNOY maintain near-constant throughput as the multiplier grows, but their recall declines, and their index build time and memory rise, making them unsuitable when datasets grow exponentially. CAKES’s tree builds orders of magnitude faster with far less memory; while queries are slower than HNSW/ANNOY, CAKES attains near-constant scaling with perfect recall.

The fastest CAKES variant depends on dataset and cardinality. On Fashion-MNIST, Glove-25, and Sift, Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as cardinality increases and consistently outperform linear search at high cardinalities. This variability motivates auto-tuning (Sec.~\ref{sec:methods:auto-tuning}); future work will better characterize when each algorithm excels and enable more sophisticated selection from dataset properties.

We stress that CAKES targets \emph{big} data: linear search is better at small sizes, but for each ANN-Benchmarks dataset tested, CAKES overtakes linear search at sufficiently large cardinality—about $10^{5}$ for Fashion-MNIST and $10^{6}$ for Glove-25. On the high-LFD Random dataset, CAKES never beats linear search. These results support that CAKES performs well on data arising from constrained generating phenomena even as cardinality grows exponentially.

@@

@@

Despite CAKES’s generality, effective deployment on real data can require domain knowledge. In protein sequence search, sequences vary widely in length (typically 30–1000 amino acids, with outliers) over a 20-letter alphabet. Numerous algorithms exist~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and their development relied on deep domain expertise.

Simple edit distance can be poor when lengths vary substantially. CAKES can get around that issue by instead using alternative metrics or preprocessing—for example, represent each sequence as a set of $k$-mers and use Jaccard distance, as in~\cite{kim2021entrance}. If the distance relies on global alignment (e.g., Needleman–Wunsch~\cite{needleman1970general}), the dataset can be partitioned into mutually exclusive length bins with a separate CAKES tree per bin; at query time, route the sequence to its bin and a small number of adjacent bins.

@@

This study suggests several directions. First, we want to extended our evaluation to more datasets and distance functions not supported by FAISS, HNSW, or ANNOY, including Wasserstein distance~\cite{vallender1974calculation} for probability distributions (particularly high-dimensional) and Tanimoto distance~\cite{bajusz2015tanimoto} for comparing molecular structures via maximal common subgraphs. Adding a new distance to CAKES requires only a Rust implementation.

Second, we intend to improve the augmentation procedure (Sec.~\ref{sec:methods:synthetic-data}) to better preserve topological structure of datasets, e.g., by favoring perturbations along the leading principal components of the local manifold.

Third, we plan to explore CAKES algorithms' use in a streaming environment, enabling online updates to the tree as points are inserted or deleted.

Finally, extend anomaly detection in CHAODA~\cite{ishaq2021clustered} by adding graph-based methods that use the distribution of distances among the $k$ nearest neighbors of cluster centers.

@@

CLAM and CAKES are implemented in Rust and released under the MIT license at https://github.com/URI-ABD/clam.
